Supplementary material for "Hack Weeks as a model for Data Science Education and Collaboration"

### 2018-05-14 conclusions.tex
Final changes

As introduced above, hack weeks serve multiple purposes, including dissemination of state-of-the-art technological advances through the scientific community, building collaborations between academic subdisciplines and fostering interdisciplinary research as well as  promoting open science and reproducibility.
Initial results from three events held in 2016 and 2017 in three different fields (astronomy, geosciences and neurosciences) indicate that hack weeks succeed at all of these objectives.

>>>>

Hack weeks serve multiple purposes, including dissemination of state-of-the-art technological advances through the scientific community, building collaborations between academic subdisciplines and fostering interdisciplinary research as well as  promoting open science and reproducibility.
Initial results from three events held in 2016 and 2017 in three different fields (astronomy, geosciences and neurosciences) indicate that hack weeks succeed at all of these objectives.

### 2018-05-14 design.tex
Minor wording changes
Final changes

A hackweek requires a flexible workspace environment with ample opportunity for re-configuration. 
Participants must have access to rooms that can accommodate lectures combined with interactive exchanges and individual work on laptops. 
Workspaces must also accommodate interactive project work where small teams can gather and work together (for more detailed considerations, see also the SM, Sections 2 and 3).    
%Universities are an obvious first choice for hosting a hack week given their access to scientific resources, research support and infrastructure. 
%However traditional university lecture halls often do not meet the hack week criteria for interactive exchanges.
Fortunately, many universities are experimenting with other types of spaces: for example, the adoption of active learning teaching methods \cite{prince2004} has led to the development of modular classrooms, designed for group activities and flexible seating arrangements.

Another important design consideration is group size (see also SM, Section 2).
If the group is too large, chances for random participant exchanges are reduced, and knowledge transfer may decline as the workshop fractures into smaller groups, often among participants who already know each other.
If the group is small, participant selection must be especially carefully managed in order to achieve the desired level of diversity among participants to foster new collaborations across sub-fields and disciplines.
In the past, we have found groups with sizes between 50 and 70 participants to work well in enabling a breadth of projects while allowing the workshop to function as a cohesive group. However, we encourage organizers to experiment with group size depending on their workshop goals.

>>>>

A hack week requires a flexible workspace environment with ample opportunity for re-configuration. 
Participants must have access to rooms that can accommodate lectures combined with interactive exchanges and individual work on laptops. 
Workspaces must also accommodate interactive project work where small teams can gather and work together (for more details, see also the SM, Sections 2 and 3).    
%Universities are an obvious first choice for hosting a hack week given their access to scientific resources, research support and infrastructure. 
%However traditional university lecture halls often do not meet the hack week criteria for interactive exchanges.
Fortunately, many universities are experimenting with other types of spaces: for example, the adoption of active learning teaching methods \cite{prince2004} has led to the development of modular classrooms, designed for group activities and flexible seating arrangements.

Another important design consideration is group size (see also SM, Section 2).
If the group is too large, chances for random participant exchanges are reduced, and knowledge transfer may decline as the workshop fractures into smaller groups, often among participants who already know each other.
If the group is small, participant selection must be especially carefully managed in order to achieve the desired level of diversity among participants to foster new collaborations.
In the past, we have found groups with sizes between 50 and 70 participants to work well in enabling a breadth of projects while allowing the workshop to function as a cohesive group. We encourage organizers to experiment with group size depending on their workshop goals.

### 2018-05-14 intro.tex
Final changes

\begin{figure}
\begin{center}
\includegraphics[width=8cm]{NewHackSpectrum.pdf}
\caption{{\bf Comparison of Extracurricular Workshop Models}: Different types of events lie on a spectrum between the emphasis on pedagogy (e.g. Software Carpentry workshop) to an emphasis on project-based/hack-based activities (e.g. at science-oriented hackathons). Hack weeks also vary in the degree of emphasis on projects (Astro hack week, which included unstructured time for group projects from the first day) or pedagogy (Neuro hack week, which included an entire first day of data science tutorials, with less time for projects)}
\label{fig:hackspectrum}
\end{center}
\end{figure}

>>>>

\begin{figure}
\begin{center}
\includegraphics[width=8cm]{NewHackSpectrum.pdf}
\caption{{\bf Comparison of Extracurricular Workshop Models}: Different types of events lie on a spectrum between the emphasis on pedagogy (e.g. Software Carpentry workshop) to an emphasis on project-based/hack-based activities (e.g. at science-oriented hackathons). Hack weeks also vary in the degree of emphasis on projects (Astro Hack Week, which included unstructured time for group projects from the first day) or pedagogy (Neuro hack week, which included an entire first day of data science tutorials, with less time for projects)}
\label{fig:hackspectrum}
\end{center}
\end{figure}

### 2018-05-14 results.tex
Final changes

\begin{table*}
%\begin{table*}[hbtp]
%\renewcommand{\arraystretch}{1.3}
\footnotesize
\caption{}
\begin{threeparttable} 
\begin{tabularx}{18cm}{p{3.4cm}p{0.3cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}}
\toprule
\textbf{Question} & \multicolumn{6}{c}{\textbf{Astro Hack Week}} & \multicolumn{6}{c}{\textbf{GeoHackWeek}} & \multicolumn{6}{c}{\textbf{NeuroHackWeek}} \\ \midrule
 & $N$ & $\mathrm{dof}_{\chi^2}^\emph{a}$ & $\chi^2$ & $p$ & $\mathrm{dof}_{\phi_c}^\emph{b}$ & $\phi_c$ & $N$ & $\mathrm{dof}_{\chi^2}^\emph{a}$ & $\chi^2$ & $p$ & $\mathrm{dof}_{\phi_c}^\emph{b}$ & $\phi_c$ & $N$ & $\mathrm{dof}_{\chi^2}^\emph{a}$ & $\chi^2$ & $p$ & $\mathrm{dof}_{\phi_c}^\emph{b}$ & $\phi_c$ \\ \midrule
\multicolumn{16}{l}{\textbf{Comparison with career stage (early-career stage; faculty and research scientists; other)}}  \\ \midrule
New topics, tools, methods (a) & 61 & 10 & 5.64 & 0.85 & 1.93 & 0 & 36 & 6 & 2.51 & 0.87 & 1.88 & 0 & 41& 10 & 15.53 & 0.15 & 1.9 & 0.23 \\
Improvement of research (b) & 61 & 6 &  9.89 & 0.13 & 1.93 & 0.18 & 35 & 4 & 12.41 & 0.02 & 1.88 & 0.35 & 40 & 6 & 7.42 & 0.28 & 1.90 & 0.13 \\
Better science (c) & 59 & 6 & 3.79 & 0.70 & 1.93 & 0 & 35 & 4 & 10.61 & 0.03 & 1.88 & 0.31 & 40 & 4 &  5.06 & 0.29 & 1.90 & 0.11\\ 
Teaching (d) & 61 & 10 & 13.65 & 0.19 & 1.93 & 0.17 & 36 & 10 & 13.68 & 0.19 & 1.88 & 0.22 & 40 & 10 & 6.90 & 0.74 & 1.90 & 0 \\ \midrule
\multicolumn{16}{l}{\textbf{Gender Identity (non-minority; minority$^\emph{c}$)}} \\ \midrule
Improvement of research (b) & 60 & 3 & 2.08 & 0.56 & 0.98 & 0 & 35 & 2 & 5.26 & 0.07 & 0.97 & 0.30 & 40 & 3 & 7.62 & 0.05 & 0.97 & 0.34\\
Teaching (d) & 61 & 5 & 7.30 & 0.20 & 0.98 & 0.19 & 36 & 5 & 12.71 & 0.26 & 0.97 & 0.46 & 40 &5 & 3.17 & 0.67 & 0.97 & 0 \\
Connection building (e) & 61 & 4 & 3.70 & 0.45 & 0.98 & 0 & 36 & 3 & 5.89 & 0.12 & 0.97 & 0.28 & 41 & 3 & 4.42 & 0.22 & 0.98 & 0.18\\
Valued contributions (f) & 57 & 4 & 5.98 & 0.20 & 0.98 & 0.18 & 34 & 5 & 9.46 & 0.09 & 0.97 & 0.36 & 38 & 5 & 6.75 & 0.34 & 0.97 & 0.21 \\ \midrule
\multicolumn{16}{l}{\textbf{Racial/ethnic Identity (white; non-white)}} \\ \midrule
Improvement of research (b)  & 59 & 3 & 0.97 & 0.81 & 0.98 & 0 & 34 &2 & 0.01 & 0.99 & 0.97 & 0 & 39 & 3 & 2.79 & 0.43 & 0.97 & 0 \\
Teaching (d) & 61 & 5 & 5.63 & 0.34 & 0.98 & 0.09 & 34 & 4 & 1.29 & 0.86 & 0.97 & 0 & 39 & 5 & 3.06 & 0.69 & 0.97 & 0 \\
Connection building (e) & 61 & 4 & 1.64 & 0.80 & 0.98 & 0 & 34 & 3 & 8.42 & 0.04 & 0.97 & 0.40 & 40 & 3 & 3.49 & 0.32 & 0.97 & 0.10 \\
Valued contributions (f) & 58 & 4 & 13.14 & 0.01 & 0.98 & 0.40 & 32 & 5 & 2.49 & 0.78 & 0.96 & 0 & 37 & 5 & 4.05 & 0.54 & 0.97 & 0 \\ 
\bottomrule
\end{tabularx}
   \begin{tablenotes}
      \item{Respondents to a post-event evaluation survey indicate their level of agreement on a six-point Likert-type scale to a range of statements (letters correspond to panels in Figure \ref{fig:survey}). Results are for hack weeks held in 2016 and 2017 for AHW and in 2016 for GHW and NHW, respectively. The number of respondents for each question is included in the table above. Additional figures showing the distributions of responses for each question are available in the SM, Section 6.}
      \item[\emph{a}]{While the range of Likert items was the same for every question, the degrees of freedom for $\chi^2$ and Cram\'{e}r's V may differ depending on the number of items with non-zero responses, and the number of non-zero items in the independent variable.}
      \item[\emph{b}]{The degrees of freedom for Cram\'{e}r's V corrected according to~\citep{bergsma2013}}
      \item[\emph{c}]{We defined ``minority'' with respect to the field of research and grouped women and non gender-conforming participants into this category.}

\end{tablenotes}

@@

Focusing on the outcomes of astro-, geo- and neuro- hack weeks (AHW, GHW, NHW, respectively) from 2016 and 2017, we find that most participants self-report successful learning outcomes (AHW: 76\%, GHW: 89\%, NHW: 79\% for responses ``somewhat agree'', ``agree'' and ``strongly agree''; Figure \ref{fig:survey}, (a)).
The overwhelming majority of respondents at the hack weeks ($>95\%$ for all three events) believed that they learned things that improve their day-to-day research, and that attendance has made them a better scientist (Figure \ref{fig:survey}, (b, c)).
%More specifically, we compared learning outcomes in data visualization (the only topic explicitly shared between all three hack weeks; Figure \ref{fig:survey}, (d, e). We asked participants to subjectively rate their knowledge before the hack week, and find skill levels to be broadly distributed. This is expected, given the goal of diversity during the selection stage. We find that most participants have positive learning outcomes at all three hack weeks for data visualization, but that outcomes vary strongly, as is expected, too, with a group that is this diverse. We also note that there might be systematic effects and biases due to the subjectivity of rating knowledge and learning outcomes.

Because peer learning is a major mode of knowledge transfer at hack weeks, we asked participants whether they taught other participants.
We find that again a majority agrees with this statement to some degree (AHW: 79\%, GHW: 69\%, NHW: 75\%; Figure \ref{fig:survey}, (d)), though responses are not as unequivocal as they are in some of the other categories.
The majority of participants felt that they built valuable connections to other researchers (Figure \ref{fig:survey}, (e)), especially at Neuro Hack Week, where more than 64\% of participants strongly agreed with this statement.

Given the broad diversity in skills and backgrounds of participants admitted to our hack weeks, a dependence of learning outcomes and teaching on career stage is plausible. At the same time, we put forth the hypothesis earlier in the this paper that peer learning is a major mode of knowledge transfer during hack weeks, and that this type of learning takes place at all career stages. However, we find no strong evidence for a significant difference between early-career researchers and senior participants for any of the three hack weeks for questions regarding learning outcomes of new tools and topics (a), improvements in day-to-day research (b) or overall improvements in science (c) (all statistical results are presented in Table \ref{tab:survey}). Only for GHW we find a weak indication that early-career researchers agree more strongly that the hack week improved their day-to-day research (p=0.02) with a large effect size of $\phi_c = 0.4$ (following~\citep{cohen1988}). 
Similarly, we find no strong indication that late-career researchers self-report a higher level of teaching at hack weeks compared to early-career researchers (d).

>>>>

\begin{table*}
%\begin{table*}[hbtp]
%\renewcommand{\arraystretch}{1.3}
\footnotesize
\caption{}
\begin{threeparttable} 
\begin{tabularx}{18cm}{p{3.4cm}p{0.3cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}p{0.33cm}}
\toprule
\textbf{Question} & \multicolumn{6}{c}{\textbf{Astro Hack Week}} & \multicolumn{6}{c}{\textbf{GeoHackWeek}} & \multicolumn{6}{c}{\textbf{NeuroHackWeek}} \\ \midrule
 & $N$ & $\mathrm{dof}_{\chi^2}^\emph{a}$ & $\chi^2$ & $p$ & $\mathrm{dof}_{\phi_c}^\emph{b}$ & $\phi_c$ & $N$ & $\mathrm{dof}_{\chi^2}^\emph{a}$ & $\chi^2$ & $p$ & $\mathrm{dof}_{\phi_c}^\emph{b}$ & $\phi_c$ & $N$ & $\mathrm{dof}_{\chi^2}^\emph{a}$ & $\chi^2$ & $p$ & $\mathrm{dof}_{\phi_c}^\emph{b}$ & $\phi_c$ \\ \midrule
\multicolumn{16}{l}{\textbf{Comparison with career stage (early-career stage; faculty and research scientists; other)}}  \\ \midrule
New topics, tools, methods (a) & 61 & 10 & 5.64 & 0.85 & 1.93 & 0 & 36 & 6 & 2.51 & 0.87 & 1.88 & 0 & 41& 10 & 15.53 & 0.15 & 1.9 & 0.23 \\
Improvement of research (b) & 61 & 6 &  9.89 & 0.13 & 1.93 & 0.18 & 35 & 4 & 12.41 & 0.02 & 1.88 & 0.35 & 40 & 6 & 7.42 & 0.28 & 1.90 & 0.13 \\
Better science (c) & 59 & 6 & 3.79 & 0.70 & 1.93 & 0 & 35 & 4 & 10.61 & 0.03 & 1.88 & 0.31 & 40 & 4 &  5.06 & 0.29 & 1.90 & 0.11\\ 
Teaching (d) & 61 & 10 & 13.65 & 0.19 & 1.93 & 0.17 & 36 & 10 & 13.68 & 0.19 & 1.88 & 0.22 & 40 & 10 & 6.90 & 0.74 & 1.90 & 0 \\ \midrule
\multicolumn{16}{l}{\textbf{Gender Identity (non-minority; minority)}} \\ \midrule
Improvement of research (b) & 60 & 3 & 2.08 & 0.56 & 0.98 & 0 & 35 & 2 & 5.26 & 0.07 & 0.97 & 0.30 & 40 & 3 & 7.62 & 0.05 & 0.97 & 0.34\\
Teaching (d) & 61 & 5 & 7.30 & 0.20 & 0.98 & 0.19 & 36 & 5 & 12.71 & 0.26 & 0.97 & 0.46 & 40 &5 & 3.17 & 0.67 & 0.97 & 0 \\
Connection building (e) & 61 & 4 & 3.70 & 0.45 & 0.98 & 0 & 36 & 3 & 5.89 & 0.12 & 0.97 & 0.28 & 41 & 3 & 4.42 & 0.22 & 0.98 & 0.18\\
Valued contributions (f) & 57 & 4 & 5.98 & 0.20 & 0.98 & 0.18 & 34 & 5 & 9.46 & 0.09 & 0.97 & 0.36 & 38 & 5 & 6.75 & 0.34 & 0.97 & 0.21 \\ \midrule
\multicolumn{16}{l}{\textbf{Racial/ethnic Identity (white; non-white)}} \\ \midrule
Improvement of research (b)  & 59 & 3 & 0.97 & 0.81 & 0.98 & 0 & 34 &2 & 0.01 & 0.99 & 0.97 & 0 & 39 & 3 & 2.79 & 0.43 & 0.97 & 0 \\
Teaching (d) & 61 & 5 & 5.63 & 0.34 & 0.98 & 0.09 & 34 & 4 & 1.29 & 0.86 & 0.97 & 0 & 39 & 5 & 3.06 & 0.69 & 0.97 & 0 \\
Connection building (e) & 61 & 4 & 1.64 & 0.80 & 0.98 & 0 & 34 & 3 & 8.42 & 0.04 & 0.97 & 0.40 & 40 & 3 & 3.49 & 0.32 & 0.97 & 0.10 \\
Valued contributions (f) & 58 & 4 & 13.14 & 0.01 & 0.98 & 0.40 & 32 & 5 & 2.49 & 0.78 & 0.96 & 0 & 37 & 5 & 4.05 & 0.54 & 0.97 & 0 \\ 
\bottomrule
\end{tabularx}
   \begin{tablenotes}
      \item{Respondents to a post-event evaluation survey indicate their level of agreement on a six-point Likert-type scale to a range of statements (letters correspond to panels in Figure \ref{fig:survey}). Results are for hack weeks held in 2016 and 2017 for AHW and in 2016 for GHW and NHW, respectively. The number of respondents for each question is included in the table above. Additional figures showing the distributions of responses for each question are available in the SM, Section 6.}
      \item[\emph{a}]{While the range of Likert items was the same for every question, the degrees of freedom for $\chi^2$ and Cram\'{e}r's V may differ depending on the number of items with non-zero responses, and the number of non-zero items in the independent variable.}
      \item[\emph{b}]{The degrees of freedom for Cram\'{e}r's V corrected according to~\citep{bergsma2013}}
\end{tablenotes}

@@

Focusing on the outcomes of astro-, geo- and neuro- hack weeks (AHW, GHW, NHW, respectively) from 2016 and 2017, we find that most participants self-report successful learning outcomes (AHW: 76\%, GHW: 89\%, NHW: 79\% for responses ``somewhat agree'', ``agree'' and ``strongly agree''; Figure \ref{fig:survey}, (a)).
The overwhelming majority of respondents at the hack weeks ($>95\%$ for all three events) believed that they learned things that improve their day-to-day research, and that attendance has made them a better scientist (Figure \ref{fig:survey}, (b, c)).
%More specifically, we compared learning outcomes in data visualization (the only topic explicitly shared between all three hack weeks; Figure \ref{fig:survey}, (d, e). We asked participants to subjectively rate their knowledge before the hack week, and find skill levels to be broadly distributed. This is expected, given the goal of diversity during the selection stage. We find that most participants have positive learning outcomes at all three hack weeks for data visualization, but that outcomes vary strongly, as is expected, too, with a group that is this diverse. We also note that there might be systematic effects and biases due to the subjectivity of rating knowledge and learning outcomes.
Because peer learning is a major mode of knowledge transfer at hack weeks, we asked participants whether they taught other participants.
We find that again a majority agrees with this statement to some degree (AHW: 79\%, GHW: 69\%, NHW: 75\%; Figure \ref{fig:survey}, (d)), though responses are not as unequivocal as they are in some of the other categories.
The majority of participants felt that they built valuable connections to other researchers (Figure \ref{fig:survey}, (e)), especially at NHW, where more than 64\% of participants strongly agreed with this statement.

Given the broad diversity in skills and backgrounds of participants admitted to our hack weeks, a dependence of learning outcomes and teaching on career stage is plausible. At the same time, we suggested earlier in this paper that peer learning is a major mode of knowledge transfer during hack weeks, and that this type of learning takes place at all career stages. We find no strong evidence for a significant difference between early-career researchers and senior participants for any of the three hack weeks for questions regarding learning outcomes of new tools and topics (a), improvements in day-to-day research (b) or overall improvements in science (c) (all statistical results are presented in Table \ref{tab:survey}). Only for GHW we find a weak indication that early-career researchers agree more strongly that the hack week improved their day-to-day research (p=0.02) with a large effect size of $\phi_c = 0.4$ (following~\citep{cohen1988}). 
Similarly, we find no strong indication that late-career researchers self-report a higher level of teaching at hack weeks compared to early-career researchers (d).

### 2018-05-14 what.tex
Final changes

Our hack weeks combine structured, tutorial-style instruction with informal education and peer learning opportunities occurring within projects and hacks. In a space spanned by pedagogical focus on one axis and focus on project work on the other, the hack weeks we have organized are designed to lie somewhere in between traditional summer schools and hackathons, where we believe they fill a space not currently fully addressed by existing models (Figure \ref{fig:hackspectrum}).

The hackathon, a time-bounded, collaborative events that bring together participants around a shared challenge or learning objective \cite{Decker2015}, forms one primary axis of our events. 
Hackathons originated from the open-source software movement and have historically focused on software and technology development. 
In recent years hackathons have evolved into a model providing opportunities for intensive, interdisciplinary collaboration \cite{Groen2015-cj} and education \cite{Kienzler2015-zu,Lamers2014-xf} in the sciences. 
Core element of all hackathons include opportunities for networking, strengthening of social ties, and the building of community connections, both within and across disciplines.
Building on these core elements hackathons have been implemented in different ways depending on the overall purpose, mode of participation, style of work environment and participant motivation \cite{Drouhard2017}. 
``Catalytic'' hackathons seek novel project ideas aimed at solving a tractable, well-defined challenge.
``Contributive'' hackathons seek to improve to an existing effort through focused work on discrete tasks, for example to make up for deficiencies in an ongoing project.
Finally, ``Communal'' hackathons place a strong focus on building a culture of practice and developing resources within an existing community, often defined by a specific domain of knowledge.

@@

Our hack weeks extend the scientifically-focused, communal hackathon model into a space that includes a strong element of pedagogy and peer learning. They aim to synthesize different goals and strategies from both models: they are much more participant-driven than a summer school, but have a stronger focus on pedagogy than a hackathon. Where a summer school is often organized around a framework of lectures and tutorials known in advance, hack weeks leave the majority of time unstructured. Tutorials at hack weeks often serve as an entry point into a topic for further exploration and learning, rather than an in-depth study. This is closely related to the nature of the material taught at these events. Data science is a new, emerging area combining knowledge and skills from many different fields, only a fraction of which can be communicated within a short tutorial. As organizers, we make a curated selection of a small number of tutorials each year, and encourage participants to use the unstructured time to self-organize tutorials on any topic as needed, often on topics vastly different to those discussed in the formal lectures.

Additionally, as we lay out in the subsequent section below, multi-directional knowledge transfer is one of many goals of a hack week, and participants often attend with more than one objective. The unstructured schedule and inherent lack of focus on specific topics within a hack week present both an opportunity and a risk. They uniquely allow organizers to tailor the content of the workshop to the needs of the participants in an ad-hoc fashion, including the number of tutorials and their content. This way, the group as a whole can respond quickly or react to unforeseen challenges and opportunities. They encourage participants to self-organize in many different forms: experts working with other experts, mentoring relationships between experts and non-experts, or study groups among non-experts, to name but a few. Hack weeks also allow participants to experiment with projects and ideas beyond their day-to-day research: our hack weeks explicitly encourage projects e.g.~ around outreach and work aimed at improving the scientific community itself.

There is, however, a major risk in the lack of focus: by wanting to be too many things at once, a hack week can run the danger of not doing any of them well. Because tutorials are not the major focus of a hack week, the knowledge gained by participants in these tutorials may not be very deep. Additionally, a hack week carries a much larger risk of failure if objectives and expectations are not set by organizers well in advance, and clearly communicated to participants, because they often require more preparation from the side of participants than other events (e.g. defining project ideas, researching possible connections to other participants, bringing along their own tutorials). 
Because of these risks, organizers face a much larger degree of uncertainty both before and during a hack week, and need to be prepared to focus much of their energy on thoughtful selection and management of participants, keeping track of the goals different attendees may have, and careful facilitation of the large range of different types of activities at any given time (see also the following sections as well as the SM, Section 4).

>>>>

Our hack weeks combine structured, tutorial-style instruction with informal education and peer learning opportunities occurring within projects and hacks. Additionally, they provide opportunities to build new collaborations and network. In a space spanned by pedagogical focus on one axis and focus on project work on the other, the hack weeks we have organized are designed to lie somewhere in between traditional summer schools and hackathons, where we believe they fill a space not currently fully addressed by existing models (Figure \ref{fig:hackspectrum}).

The hackathon, a time-bounded, collaborative events that bring together participants around a shared challenge or learning objective \cite{Decker2015}, forms one primary axis of our events. 
Hackathons originated from the open-source software movement and have historically focused on software and technology development. 
In recent years hackathons have evolved into a model providing opportunities for intensive, interdisciplinary collaboration \cite{Groen2015-cj} and education \cite{Kienzler2015-zu,Lamers2014-xf} in the sciences. 
Core element of all hackathons include opportunities for networking, strengthening of social ties, and the building of community connections, both within and across disciplines.
Building on these core elements hackathons have been implemented in different ways depending on the overall purpose, mode of participation, style of work environment and participant motivation \cite{Drouhard2017}. 
%``Catalytic'' hackathons seek novel project ideas aimed at solving a tractable, well-defined challenge.
%``Contributive'' hackathons seek to improve to an existing effort through focused work on discrete tasks, for example to make up for deficiencies in an ongoing project.
%Finally, ``Communal'' hackathons place a strong focus on building a culture of practice and developing resources within an existing community, often defined by a specific domain of knowledge.

@@

Our hack weeks extend the scientifically-focused, communal hackathon model into a space that includes a strong element of pedagogy and peer learning. They aim to synthesize different goals and strategies from both models: they are much more participant-driven than a summer school, but have a stronger focus on pedagogy than a hackathon. Where a summer school is often organized around a framework of lectures and tutorials known in advance, hack weeks leave the majority of time unstructured. Tutorials at hack weeks often serve as an entry point into a topic for further exploration and learning, rather than an in-depth study. This is closely related to the nature of the material taught at these events. Data science is a new, emerging area combining knowledge and skills from many different fields, only a fraction of which can be communicated within a short tutorial. As organizers, we make a curated selection of a small number of tutorials each year, and encourage participants to use the unstructured time to self-organize tutorials on any topic as needed.%, often on topics vastly different to those discussed in the formal lectures.

The unstructured schedule and inherent lack of focus on specific topics within a hack week present both an opportunity and a risk. They uniquely allow organizers to tailor the content of the workshop to the needs of the participants in an ad-hoc fashion, including the number of tutorials and their content. This way, the group as a whole can respond quickly or react to unforeseen challenges and opportunities. They encourage participants to self-organize in many different forms: experts working with other experts, mentoring relationships between experts and non-experts, or study groups among non-experts, to name but a few. Hack weeks also allow participants to experiment with projects and ideas beyond their day-to-day research: our hack weeks explicitly encourage projects e.g.~ around outreach and work aimed at improving the scientific community itself.

There is, however, a major risk in the lack of focus: by wanting to be too many things at once, a hack week can run the danger of not doing any of them well. Because tutorials are not the major focus of a hack week, the knowledge gained by participants in these tutorials may not be very deep. Additionally, a hack week carries a much larger risk of failure if objectives and expectations are not set by organizers well in advance, and clearly communicated to participants, because they often require more preparation from the side of participants than other events (e.g. define project ideas, research possible connections to other participants). 
Because of these risks, organizers face a much larger degree of uncertainty both before and during a hack week, and need to be prepared to focus much of their energy on thoughtful selection and management of participants, keeping track of the goals different attendees may have, and careful facilitation of the large range of different types of activities at any given time (see also the following sections as well as the SM, Section 4).
